Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-19 paper/main.tex
restructure introductory text

%###################################################################################################
\section{Introduction}
\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells \textbf{TODO CITATION}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

Despite their popularity, \acp{abm} face multiple challenges in practise.
The construction of a fully functional \ac{abm} simulation from the ground up constitutes an
intricate and laborious process, necessitating a comprehensive array of skills. It is therefore
common practise to reuse existing simulation tools, which, in turn, brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability \textbf{TODO CITATION}.
Typically researchers estimate the parameter of their model using methods such as Maximum Likelihood
or Maximum a Posteriori~\cite{Gbor2015,Banga2008,Ashyraliyev2009}, check for parameter
identifiability using methods like profile-likelihood or apply model reduction
techniques~\cite{Kreutz2013,Raue2014,Simpson2025}.

In the context of \acp{abm}, however, it is often only shown that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.
In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria as an example.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model. We deliberately distinguish between conceptually separate components within
this framework.

Mechanical properties of rod-shaped bacteria have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015,Balagam2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
Rod-shaped Bacteria are an interesting target system since their physical form provides more surface
information compared to other forms such as spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

@@

\textbf{
    Was mir hier nicht gefällt, ist, dass du von Parameter schätzen bei ABMs kommend ziemlich
    detailliert über Mechanik redest.
    Das passt nicht.
    Man erwartet mehr Infos über den Blueprint und ist verwirrt.
    Worum geht es hier?
}
\textit{Ich habe den blueprint vollkommen entfernt. Er macht kein Sinn im MS. Der Rest sollte noch
angepasst werden.}

This work addresses some of these limitations.
We use microscopic image time series as inputs to obtain information about the spatial configuration
of the cells.
This information can be used as initial values for a numerical simulation.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.

>>>>

%###################################################################################################
\section{Introduction}
\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells~\cite{DeRybel2014}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

Despite their popularity, \acp{abm} face multiple challenges in practise.
The construction of a fully functional \ac{abm} simulation from the ground up constitutes an
intricate and laborious process, necessitating a comprehensive array of skills. It is therefore
common practise to reuse existing simulation tools, which, in turn, brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability~\cite{Pleyer2025Flexibility}.

Typically researchers estimate the parameters of their model using methods such as Maximum Likelihood
or Maximum a Posteriori~\cite{Gbor2015,Banga2008,Ashyraliyev2009}, check for parameter
identifiability using methods like profile-likelihood or apply model reduction
techniques~\cite{Kreutz2013,Raue2014,Simpson2025}.
However, in the context of \acp{abm} this practice is almost never followed and instead only shown
that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
This lack of systematic analysis of the involved parameters leaves a large gap of explainability of
the underlying models.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.

In this work, we showcase how a systematic analysis of parameters can be performed using a
general-purpose mechanical model for rod-shaped bacteria.
Rod-shaped Bacteria present an interesting target system since their physical form provides more
surface information compared to other forms such as spherical cells.
Their mechanical properties have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015,Balagam2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
In addition, parameters such as growth rates have been found to vary within a given cell
population~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity when
constructing models on the single-cell level.

@@

This work addresses some of these previously listed limitations.
We use microscopic image time series as inputs to obtain information about the spatial configuration
of the cells.
This information can be used as initial values for a numerical simulation and also serves as ground
truth.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.
We deliberately distinguish between conceptually separate components within this framework to allow
for downstream reusability.
The developed computational methods can be combined to systematically estimate the parameters of our
model and are readily available as a Python package.
